### 2013-07-24 doc/asplos14/evaluation.tex
insert new BFS numbers; expand on discussion

In order to put \Grappa's performance into a general context, we compare it
with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
different hardware platform, we also compare \Grappa with hand-tuned MPI
versions of BFS and GUPS running on the same hardware. Finally, we also
compare it with UTS written for UPC. We run all experiments with 64 nodes.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c|c}
&	\Grappa & XMT	&        MPI	&   UPC \\ \hline
GUPS	& 1	& 2.23	& 0.11& 	-- \\ 
BFS	& 1 &	4.33	& 14.37 & 	-- \\ 
Intsort	& 1	& 3.77 &	5.88	& -- \\
UTS (T1)	& 1	& 0.38	& -- &	0.09 \\ 
Pagerank	& 1	& 4.35	& 4.87 &	-- \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT, hand-tuned MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

\TODO{addressing reviwer2: sweep of irregularness in one aggregate or multiple dimensions
    and how a specific aspect of performance changes in different
    systems}

Table~\ref{tab:grappa-comparisons} shows the results. Overall, \Grappa is
within striking distance of XMT and well beyond UPC performance (UTS only).
However, it underperforms compared to hand-tuned MPI\@. We believe there are two reasons for this:
First, Grappa is young; much optimization and tuning of the runtime
and the application is yet to be done. Second, the MPI implementations
we used employ algorithmic-specific optimizations -- some of which
resemble what \Grappa does automatically, like message aggregation.
Grappa's GUPS results presented earlier suggest there is significant
room for improvement.

>>>>

In order to put \Grappa's performance into a general context, we compare it
with XMT running BFS, PageRank, IntSort, GUPS and UTS. Since XMT is a
different hardware platform, we also compare \Grappa with optimized MPI
versions of BFS and GUPS running on the same hardware. Finally, we also
compare it with UTS written for UPC. We run all experiments with 64 nodes and 16 cores per node.

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{l|c|c|c|c}
         & \Grappa & XMT   & MPI  & UPC \\ \hline
GUPS     & 1       & 2.23  & 0.11 & -- \\ 
BFS      & 1       & 1.63  & 3.52 & -- \\ 
Intsort  & 1       & 3.77  & 5.88 & -- \\
UTS (T1) & 1       & 0.38  & --   & 0.09 \\ 
Pagerank & 1       & 4.35  & 4.87 & -- \\ 
\end{tabular}
\end{center}
\caption{Comparing \Grappa with XMT, optimized MPI and UPC. Numbers are presented as throughput on 64 nodes normalized to \Grappa.}
\label{tab:grappa-comparisons}
\end{table}

Table~\ref{tab:grappa-comparisons} shows the results. \Grappa is able
to provide performance approaching and sometimes exceeding that of the
other systems. This data shows that, while \Grappa is good at the
operations for which it was designed, there is a cost to providing
\Grappa's generality. In general, the benchmarks whose MPI
implementations are faster than \Grappa include a implementation of a
subset of \Grappa's functionality specialized for their particular
problem. We now discuss each benchmark in more detail.

\paragraph{UTS.}
It is perhaps not surprising that the XMT is faster than
\Grappa on many of the benchmarks; after all, the XMT is custom
hardware optimized for irregular applications. The XMT can
context-switch every cycle, while \Grappa takes $~$50 ns. The XMT can
complete 100 million random-access remote reads per second per node in
hardware, while \Grappa must spend many instructions managing and
aggregating each of these reads in software.

What is more surprising is that Grappa is able to exceed the XMT's
performance on UTS. This is a demonstration of the inflexibility of
custom hardware. Recall that the UTS benchmark searches an unbalanced
tree and spawns a data-dependent number of child tasks at each
vertex. The XMT's spawn/join semantics, whose implementation is spread
across the hardware, OS, and runtime, require a task visiting a vertex
to block until all its child tasks have completed, even though the
benchmark does not require those tasks to have completed until the end
of the tree search. Furthermore, each new task is immediately made
available to all other processors in the system, even though it is
likely that processor that spawned a task will end up executing it.

Since \Grappa's tasks are implemented entirely in software, it is easy
to support arbitrary synchronization semantics. \Grappa allows both
XMT-like per-vertex task joins as well as whole-problem joins that are
a better fit for UTS. Furthermore, \Grappa's work-stealing distributed
task queue and recursive loop decomposition encourage locality in task
execution, so that the system is optimized for the common case of a
processor executing a task it spawned.

The UPC UTS implementation includes an implementation of work-stealing
but has limited ability to overlap communication with
computation. Furthermore, it must execute heavyweight synchronization
operations when stealing. \Grappa benefits from its lightweight
context switching and support for fine-grained synchronization.

\paragraph{BFS.}
The Grappa and MPI BFS implementations are quite similar. They both
use the same graph representation, and they both depend on aggregation
for performance. The difference is that the BFS' aggregation is
integrated with the traversal code and specialized for the problem:
since the only messages required are edge traversals, the aggregation
buffers store only pairs of 8-byte vertex IDs.  In contrast, \Grappa
pays execution and space overhead in order to support aggregation of
arbitrary closures; the messages \Grappa aggregates include the two
8-byte vertex IDs along with 8 bytes of deserialization information as
well as 8 bytes of synchronization information. Furthermore, the MPI
implementation writes the edge information directly into the
aggregation buffer while \Grappa executes additional code to serialize
and deserialize the messages. \TODO{Include message counts, size, and
  rates; code size too?}.

\paragraph{Other benchmarks.}

The MPI GUPS implementation includes an implementation of aggregation,
but it is not optimized to work when the aggregated data exceeds
cache, and it has limited support for concurrent communication with
multiple destinations. \Grappa benefits from being optimized for both
of these cases.

The MPI Pagerank implementation it built on top of the
highly-optimized sparse matrix support in the Trilinos~\cite{trilinos}
library. In contrast, the \Grappa implementation is fairly
straightforward.

\TODO{Describe Intsort}

\paragraph{Summary.}

Overall, \Grappa provides a general programming model at a moderate
performance cost. While \Grappa's core functionality performs well,
applications can specialize similar functionality for their problems
and obtain better performance than \Grappa on the same hardware. This
is, however, not the end of the story for \Grappa's performance;
\Grappa is a young library and, as discussed in the next section, is
limited by its implentation rather than the hardware on which it
runs. We believe there are opportunities for optimization that will
improve its performance.
